### 2019-07-29 Thesis/pages/lit_review.tex
Changes; - Added more content to the background - Merged background and literature review

Fingerprints are used to compare the similarity of two keys quickly. The comparison occurs between two digests of a secure one-way hash function. Historically these have been represented as a hexadecimal string whereon verification fingerprints are compared between two substrates, for example, a monitor screen and a business card.
This process is known as the \textit{``authentication ceremony.''}

Prior research has shown that the average human can only hold around 7-digits worth of data in their working memory\cite{miller1956magical}. This rules out the possibility of comparing complete digests. For example, SHA-1 is 40 hex digits (160-bit) and, therefore, difficult to effectively compare. Therefore, if human interaction is required; there is a need for schemes that work effectively with consideration to individual limitations.

@@

\subsection*{Definition of terms}
Throughout the paper, various terms are used to define elements of the authentication ceremony relevant to fingerprints verification. Explicit descriptions are, therefore, stated in this section. These are used to remove ambiguity and underlying associations readers may have for the words chosen.
\\\\
\textbf{\textit{Encoding schemes}} - This is the physical method of encoding used to represent the fingerprint. For example, ``Hexadecimal'' or ``Words'' are a unique way to represent a fingerprint and thus, are encoding schemes.
\\\\
\textbf{\textit{Method of comparison}} - This is how the user assesses the encoding scheme. The most common example and the one used as default is ``Compare-and-Confirm'' (CaC). CaC, as the name suggests, is the \textit{comparison} of two fingerprints on different devices or mediums that are then \textit{confirmed}. The following sections explain alternative methods of comparisons.

\subsection{Authentication ceremony performance}

\subsection*{Encoding scheme performance}
Results 
from the literature consistently show the effectiveness of language-based encodings such as Words or Sentences with accuracies ranging from 94\% \cite{dechand2016empirical}\cite{tan2017can}\cite{kainda2009usability}. In all cases, these were the best schemes from the sets assessed. The exception to this is the work performed by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study} in 2009 with Words achieving an abnormal accuracy of 63.00\%.
\\\\
Aside from textual representations were graphical schemes. Examples of schemes assessed were: Random Art\cite{perrig1999hash}, Flag\cite{ellison2003public}, T-Flag\cite{lin2010spate}, Vash\footnote{https://github.com/thevash/vash}, OpenSSH visual host key and Unicorns\footnote{https://unicornify.pictures/}, among others. These schemes had mixed accuracy with ranges as large as 50\% - 94\% in work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study}. The only other paper assessing graphical representations was the work of \textbf{Tan \textit{et al.}}\cite{tan2017can} where they also achieved mixed results with accuracies ranging from 46\% to 90\%.
\\
\begin{table}[h!]
    \makebox[\textwidth][c]{
        \input{tables/fingerprint_times.tex}
    }%
    \caption{Timing results in seconds for the related schemes}
    \label{tab:times}
\end{table}
\\
In terms of usability of graphical schemes, the literature concurred on their high usability. The comparison speed of these schemes were all among the quickest (See Table \ref{tab:times} for an overview of timings). Other work also indirectly concurred with graphical encodings having significantly quicker comparison times compared to non-graphical schemes \cite{dechand2016empirical}\cite{kainda2009usability}.
In terms of research into the performance of graphical schemes, the literature does not contain an extensive review with literature only containing two papers. There is also no overlap in the schemes assessed with each paper reviewing a unique set. This is, therefore, a promising candidate for further research.

@@

\subsection*{Methods of comparison performance}
Aside 
from ``Compare-and-Confirm'' (CaC), there is ``Compare-and-Select'' (CaS) and ``Compare-and-Enter'' (CaE). CaS is the method where one device displays the fingerprint, and the other user is provided with several options. The user then has to choose the correct value from the list of candidates. If there is no match, the user must deny the connection attempt. The creation of CaS was due to concerns that CaC would be ``too easy'' for users leading to complacency and errors\cite{uzun2007usability}. CaE is designed for scenarios where both devices might not have a display, i.e. pairing between a phone and a keyboard. One device displays the checksum. This checksum is entered into the other device. The first device then compares the entered string and checks for a match.

@@

\subsection*{Experimentation methodology comparison}
To further look into the validity of previously discussed results, it is necessary to asses how the respective studies reached these conclusions. Areas for consideration are scheme entropy, attacker strength and participant attributes.

@@

%TODO: Maybe talk about the effect of visual or mental impairments 
\subsubsection{Topic conclusion}
Overall, this review has identified several key areas suitable for further work.
The first is the performance assessment of graphical encoding schemes. Recreation of pre-existing results from \cite{hsiao2009study}\cite{tan2017can} is required to corroborate current conclusions and validate results.
Another gap in the research is the consideration into the utilization of encoding schemes in realistic conditions, i.e. ``remove vs. proximity." This topic was initially covered by \cite{shirvanian2017pitfalls} but their scope was limited. Further work, therefore, could increased the scope and touch upon a large number of schemes in these settings.
The final aspect for further work is the limited consideration into the feasibility of attacks on encoding schemes. All of the papers assessed simulated attacks and had minimal consideration for the execution of these attacks. Therefore, further work could delve into the implementation of such attacks and their feasibility in terms of compute and storage complexity.

\subsection{Encoding schemes}
Another area of research is investigations into the actual physical encodings of the hash digest. This section will briefly discuss the current research available on the creation and security of actual encoding schemes. The actual details of the operation of the schemes are outside the scope of this literature review, therefore, minimal attention will be allocated to these details.

@@

Aside from this research, there have been further informal implementations of fingerprint encodings. The first being by \textbf{Michael Rogers}\footnote{https://github.com/akwizgran/basic-english}. Rogers' implementation is a program designed to map fingerprints to pseudo-random poems. This implementation was again, empirically evaluated by Dechand \textit{et al.}\cite{dechand2016empirical}. Older work by \textbf{N. Haller} with the S/KEY\cite{haller1995s} shows the implementation of a system designed to represent a hash as a series of six short words. However, this system is designed for a one-time-password purpose and only provides word mappings for basic human usability of the password and not within a fingerprint verification context. Therefore, the wordlist has not been designed with pronounceability in mind.
\\
A very recent implementation of a word list can be found in Pretty Easy Privacy (pEp) implementation of TrustWords\footnote{https://tools.ietf.org/html/draft-birk-pep-trustwords-03}. The unique aspect of TrustWords is its mapping of a single word to 16-bits. In comparison to other literature, this is the highest number of bits-per-word seen. Full mappings (no duplication of words) would, therefore, require $2^{16}$ words in the dictionary and arguably is higher than most users vocabulary. this deviation from the norm has not been currently backed up by research. Moreover the main RFC documentation still remains in a draft stage and states \textit{``It is for further study, what minimal number of words (or entropy) should be required.''}. These aspects clearly highlight on a gap in the current literature.

\subsubsection{Topic conclusion}

@@

\subsection{Attacks on encoding schemes}
This area of research studies ways to physically execute attacks on fingerprint encoding schemes. This differs from previously examined work due to papers discussing the performance and fallibility of encoding schemes simulated the attack without consideration for how the attack would be performed. Research in this area is scant, with lots of research attention being directed towards the security of Man-in-the-Middle (MITM) attacks and not the encoding schemes themselves.

@@

\subsubsection{Topic Conclusion}
Overall the literature for this subtopic remains sparse and incomplete. Further suggested work could look into the feasibility of generating partial collisions for all textual representations alongside quantified effectiveness on users. With the possibility to concentrate on a few selected implementations. The work would aim to focus on the various physical methods used and their feasibility. This is one area the previous literature has failed to cover and has only theoretically quantified attacker strength without consideration for the actual real-world cost of these attacks.

>>>>

The chapter explains aspects required to fully comprehend the project's achievements alongside an evaluation of current literature and the respective gaps.

\section{Public-key Cryptography and Key Exchange}
Asymmetric cryptography facilitates the secure encryption of messages in end-to-end encryption (E2EE), verification of digital signatures and sharing of pre-communication secrets, among others. The use-case of E2EE messages will be the primary focus of this paper.

The asymmetry stems from the use of "Public" and "Private" keys. The Public key is used to encrypt data that only the respective Private key can decrypt. This means the keys required to communicate can be easily shared across insecure channels.

To construct an E2EE connection the initial stage will be a key exchange using the key types discussed previously. This is the sharing of a pre-shared secret between two verified parties, this will then be used to encrypt subsequent messages with symmetric encryption\footnote{This is due to the speed increase of symmetrical over asymmetrical encryption}. This, however, hinges on the verification of the initial parties, if one party impersonates another and gets to view the pre-shared secret all communication becomes decryptable. Therefore, the correct identification of parties is crucial to maintaining secure communication channels. The exploitation of this is known as a Man-in-the-middle (MiTM) attack. This can be bidirectional where the attacker can sit in the middle of a communication channel and decrypt all correspondence. Figure \ref{fig:mitm} contains a visual representation of this attack.

\begin{center}
    \input{diagrams/MITM/basic}
    \begin{figure}[h]
        \caption{Photo depicting a MITM attack}
        \label{fig:mitm}
    \end{figure}
\end{center}

Due to the assumption that the attacking party (Eve) does not have the same public-private key pair as either party (Alice or Bob) the fingerprint of the public key can be used for identification. There are, however, issues in how a user should link a key's fingerprint to real world entity. This paper will, however, assume that users know the respective parties' correct fingerprint, as the protocols used to achieve this are outside the project's scope.

The fingerprint of the key is generated by running the main components through a secure one way hash function\footnote{Explain this?}. This process produces a digest of fixed length that can be used to compare keys. Therefore, the comparison of expected and actual fingerprints is used to detect MiTM attacks and prevent impersonation.
Historically these have been represented as a hexadecimal string whereon verification fingerprints are compared between two substrates, for example, a monitor screen and a business card.
This process is known as the \textit{``authentication ceremony.''}. However, other systems are moving away from that and creating a single key from each sides' public key. In the case of WhatsApp for example, the connection fingerprint of two parties keys is the first 30 bytes of a SHA-512\footnote{5200 iterations} hash of each parties' identity key, these are then both concatenated together to form a single key\cite{whatsapp2017paper}. This produces a unique key communication pair.
\\\\
\textbf{TODO:} Explain why fingerprints are changing to one-per-pair?

Due to the manual comparison of the fingerprint, length and format is a consideration. Prior research has shown that the average human can only hold around 7-digits worth of data in their working memory\cite{miller1956magical}. This rules out the possibility of comparing complete digests. For example, SHA-1 is 40 hex digits (160-bit) and, therefore, difficult to effectively compare. Therefore, if human interaction is required; there is a need for schemes that work effectively with consideration to individual limitations.

@@

\section*{Definition of terms}
Throughout the paper, various terms are used to define elements of the authentication ceremony relevant to fingerprints verification. Explicit descriptions are, therefore, stated in this section. These are used to remove ambiguity and underlying associations readers may have for the words chosen.
\\\\
\textbf{\textit{Encoding schemes}} - This is the physical method of encoding used to represent the fingerprint. For example, ``Hexadecimal'' or ``Words'' are a unique way to represent a fingerprint and thus, are encoding schemes.
\\\\
\textbf{\textit{Method of comparison}} - This is how the user assesses the encoding scheme. The most common example and the one used as default is ``Compare-and-Confirm'' (CaC). CaC, as the name suggests, is the \textit{comparison} of two fingerprints on different devices or mediums that are then \textit{confirmed}. The following sections explain alternative methods of comparisons.

\section{Authentication ceremony performance}

\section*{Encoding scheme performance}
Results 
from the literature consistently show the effectiveness of language-based encodings such as Words or Sentences with accuracies ranging from 94\% \cite{dechand2016empirical}\cite{tan2017can}\cite{kainda2009usability}. In all cases, these were the best schemes from the sets assessed. The exception to this is the work performed by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study} in 2009 with Words achieving an abnormal accuracy of 63.00\%.
\\\\
Aside from textual representations were graphical schemes. Examples of schemes assessed were: Random Art\cite{perrig1999hash}, Flag\cite{ellison2003public}, T-Flag\cite{lin2010spate}, Vash\footnote{https://github.com/thevash/vash}, OpenSSH visual host key and Unicorns\footnote{https://unicornify.pictures/}, among others. These schemes had mixed accuracy with ranges as large as 50\% - 94\% in work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study}. The only other paper assessing graphical representations was the work of \textbf{Tan \textit{et al.}}\cite{tan2017can} where they also achieved mixed results with accuracies ranging from 46\% to 90\%.
\\
\begin{table}[h!]
    \makebox[\textwidth][c]{
        \input{tables/fingerprint_times.tex}
    }%
    \caption{Timing results in seconds for the related schemes}
    \label{tab:times}
\end{table}
\\
In terms of usability of graphical schemes, the literature concurred on their high usability. The comparison speed of these schemes were all among the quickest (See Table \ref{tab:times} for an overview of timings). Other work also indirectly concurred with graphical encodings having significantly quicker comparison times compared to non-graphical schemes \cite{dechand2016empirical}\cite{kainda2009usability}.
In terms of research into the performance of graphical schemes, the literature does not contain an extensive review with literature only containing two papers. There is also no overlap in the schemes assessed with each paper reviewing a unique set. This is, therefore, a promising candidate for further research.

@@

\section*{Methods of comparison performance}
Aside 
from ``Compare-and-Confirm'' (CaC), there is ``Compare-and-Select'' (CaS) and ``Compare-and-Enter'' (CaE). CaS is the method where one device displays the fingerprint, and the other user is provided with several options. The user then has to choose the correct value from the list of candidates. If there is no match, the user must deny the connection attempt. The creation of CaS was due to concerns that CaC would be ``too easy'' for users leading to complacency and errors\cite{uzun2007usability}. CaE is designed for scenarios where both devices might not have a display, i.e. pairing between a phone and a keyboard. One device displays the checksum. This checksum is entered into the other device. The first device then compares the entered string and checks for a match.

@@

\section*{Experimentation methodology comparison}
To further look into the validity of previously discussed results, it is necessary to asses how the respective studies reached these conclusions. Areas for consideration are scheme entropy, attacker strength and participant attributes.

@@

%TODO: Maybe talk about the effect of visual or mental impairments 
\subsection{Topic conclusion}
Overall, this review has identified several key areas suitable for further work.
The first is the performance assessment of graphical encoding schemes. Recreation of pre-existing results from \cite{hsiao2009study}\cite{tan2017can} is required to corroborate current conclusions and validate results.
Another gap in the research is the consideration into the utilization of encoding schemes in realistic conditions, i.e. ``remove vs. proximity." This topic was initially covered by \cite{shirvanian2017pitfalls} but their scope was limited. Further work, therefore, could increased the scope and touch upon a large number of schemes in these settings.
The final aspect for further work is the limited consideration into the feasibility of attacks on encoding schemes. All of the papers assessed simulated attacks and had minimal consideration for the execution of these attacks. Therefore, further work could delve into the implementation of such attacks and their feasibility in terms of compute and storage complexity.

\section{Encoding schemes}
Another area of research is investigations into the actual physical encodings of the hash digest. This section will briefly discuss the current research available on the creation and security of actual encoding schemes. The actual details of the operation of the schemes are outside the scope of this literature review, therefore, minimal attention will be allocated to these details.

@@

Aside from this research, there have been further informal implementations of fingerprint encodings. The first being by \textbf{Michael Rogers}\footnote{https://github.com/akwizgran/basic-english}. Rogers' implementation is a program designed to map fingerprints to pseudo-random poems. This implementation was again, empirically evaluated by Dechand \textit{et al.}\cite{dechand2016empirical}. Older work by \textbf{N. Haller} with the S/KEY\cite{haller1995s} shows the implementation of a system designed to represent a hash as a series of six short words. However, this system is designed for a one-time-password purpose and only provides word mappings for basic human usability of the password and not within a fingerprint verification context. Therefore, the wordlist has not been designed with pronounceability in mind.
\\\\
A very recent implementation of a word list can be found in Pretty Easy Privacy (\pep) implementation of TrustWords\footnote{https://tools.ietf.org/html/draft-birk-pep-trustwords-03}. \pep is a data encryption system that utilises PGP encryption to provide E2EE on all common channels of communication such as email of SMS. The imbedded design principles state that above all the systems should be easy to install, use and understand.

\begin{wrapfigure}{r}{6cm}
    \centering
    \includegraphics[scale=0.15]{trustwords/trustword_handshake.jpg}
    \caption{Trustword fingerprint verification}
    \label{fig:trustwords}
\end{wrapfigure}

\pep deals with the threat of MiTM attacks by having users compare the respective key fingerprints encoded as a set of words. Figure \ref{fig:trustwords} shows the \pep Android implementation of trustwords. The user pair of users then authenticate the words on an OOB (Out of Band) channel such as a phone call or in-person communication. If both users decide the words match they will accept or decline respectively.

The unique aspect of TrustWords is its mapping of a single word to 16-bits. In comparison to other literature, this is the highest number of bits-per-word seen. Full mappings (no duplication of words) would, therefore, require $2^{16}$ words in the dictionary and arguably is higher than most users vocabulary. this deviation from the norm has not been currently backed up by research. Moreover the main RFC documentation still remains in a draft stage and states \textit{``It is for further study, what minimal number of words (or entropy) should be required.''}. These aspects clearly highlight on a gap in the current literature.

\textbf{TODO: } Move pep stuff here!

\subsection{Topic conclusion}

@@

\section{Attacks on encoding schemes}
This area of research studies ways to physically execute attacks on fingerprint encoding schemes. This differs from previously examined work due to papers discussing the performance and fallibility of encoding schemes simulated the attack without consideration for how the attack would be performed. Research in this area is scant, with lots of research attention being directed towards the security of Man-in-the-Middle (MITM) attacks and not the encoding schemes themselves.

@@

\subsection{Topic Conclusion}
Overall the literature for this subtopic remains sparse and incomplete. Further suggested work could look into the feasibility of generating partial collisions for all textual representations alongside quantified effectiveness on users. With the possibility to concentrate on a few selected implementations. The work would aim to focus on the various physical methods used and their feasibility. This is one area the previous literature has failed to cover and has only theoretically quantified attacker strength without consideration for the actual real-world cost of these attacks.
